\Large Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-11-07 paper/main.tex
added text related to temporal correction, incorporated refs to barplot figure, converted barplot figure to a sidewaysfigure so that it's readable

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage[sort&compress]{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage{amsopn}

@@

We developed a permutation-based procedure to factor out the effects of
temporal clustering from the clustering scores for each feature dimension. For
a given set of recalled items (whose presentation positions are given by $x_1,
x_2, x_3, …, x_N$), we circularly shift the presentation positions by a
randomly chosen amount (between 1 and the list length) to obtain a new set of
items. Since the new set of items will have the same (average) temporal
distances between successive recalls, the temporal clustering score for the new
set of items is equal (on average) to the temporal clustering score for the
original recalls. However, we can then re-compute the feature clustering score
for those new items. Finally, we can compute a ``temporally corrected'' feature
clustering score by computing the average percentile rank of the observed (raw)
feature clustering score within the distribution of circularly shifted feature
clustering scores. This new temporally corrected score provides an estimate of
the observed degree of feature clustering over and above what could be
accounted for by temporal clustering alone.

@@

= 0.067$). Participants' performances on early lists in all of the other order
manipulation conditions were indistinguishable from performance on the early
feature rich lists ($\|t\|$s $< 1.013, p$s $> 0.314$). Participants in both of
the semantically ordered conditions exhibited stronger temporal clustering on
early lists (versus early feature rich lists; category: $t(95) = 8.508, p <
0.001$; size: $t(95) = 2.429, p = 0.017$). Participants in the length-ordered
condition tended to exhibit \textit{less} temporal clustering on early lists
relative to early feature rich lists ($t(95) = -1.666, p = 0.099$), whereas
participants in the first letter-ordered condition exhibited stronger temporal
clustering on early lists ($t(95) = 2.587, p = 0.011$). Participants in the
visually ordered conditions exhibited more similar performance on early lists,
relative to early feature rich lists (color: $t(96) = -1.064, p = 0.290$; we
found a trending enhancement for participants in the location-ordered
condition: $t(95) = 1.682, p = 0.096$). We also compared feature-based
clustering on early lists across the order manipulation and feature rich
conditions. Since these results were similar across both semantic conditions
(category and size), both lexicographic conditions (length and first letter),
and both visual conditions (color and location), here we aggregate data from
conditions that manipulated each of these three feature groupings in our
comparisons, to simplify the presentation. On early lists, participants in the
semantically ordered conditions exhibited stronger semantic clustering relative
to participants in the feature rich condition (category: $t(125) = 2.524, p =
0.013$; size: $t(125) = 3.510, p = 0.001$), but showed no reliable differences
in lexicographic (length: $t(125) = 0.539, p = 0.591$; first letter: $t(125) =
-0.587, p = 0.558$) or visual (color: $t(125) = -0.579, p = 0.564$; location:
$t(125) = -0.346, p = 0.730$) clustering. Similarly, participants in the
lexicographically ordered conditions exhibited stronger (relative to feature
rich participants) lexicographic clustering (length: $t(125) = 3.426, p =
0.001$; first letter: $t(125) = 3.236, p = 0.002$) on early lists, but showed
no reliable differences in semantic (category: $t(125) = -1.078, p = 0.283$;
size: $t(125) = -0.310, p = 0.757$) or visual (color: $t(125) = -0.209, p =
0.835$; location: $t(125) = -0.004, p = 0.997$) clustering. And participants in
the visually ordered conditions exhibited stronger visual clustering (again,
relative to feature rich participants, and on early lists; color: $t(126) =
2.099, p = 0.038$; location: $t(126) = 4.392, p < 0.001$), but showed no
reliable differences in semantic (category: $t(126) = 0.204, p = 0.839$; size:
$t(126) = -0.093, p = 0.926$) or lexicographic (length: $t(126) = 0.714, p =
0.476$; first letter: $t(126) = 0.820, p = 0.414$) clustering. Taken together,
these order manipulation results suggest several broad patterns
(Figs.~\ref{fig:recall-dynamics}A, \ref{fig:fingerprints}). First, most of the
order manipulations we carried out did \textit{not} reliably affect overall
recall performance. Second, most of the order manipulations increased
participants' tendencies to temporally cluster their recalls. Third, all of the
order manipulations enhanced participants' clustering of each condition's
target feature (i.e., semantic manipulations enhanced semantic clustering,
lexicographic manipulations enhanced lexicographic clustering, and visual
manipulations enhanced visual clustering) while leaving clustering along other

@@

affect lexicographic or visual clustering, and so on).

@@

\begin{figure}[tp] \centering
    \includegraphics[width=\textwidth]{figures/memory_perf_barchart_compare}

\caption{\textbf{Recall probability and clustering scores on early and late
lists.} The bar heights display the average (across participants) recall
probabilities (\textbf{A.}), temporal clustering scores (\textbf{B.}), feature
clustering scores (\textbf{C.}), and residual feature clustering scores (after
factoring out temporal clustering effects; \textbf{D.}) for early (gray) and
late (gold) lists. For the feature rich bars (left), the feature clustering
scores are averaged across features. For the order manipulation conditions,
feature clustering scores are displayed for the focused-on feature for each
condition (e.g., category clustering scores are displayed for the category
condition, and so on). All panels: error bars denote bootstrap-estimated 95\%
confidence intervals. The horizontal dotted lines denote the average values
(across all lists and participants) for the feature rich condition. The bars
denote $t$-tests between the corresponding bars, and the asterisks denote the
Benjamini-Hochberg-corrected $p$-values. Comparisons for which corrected $p
\geq 0.05$ are not shown.} \label{fig:barplots}

>>>>

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage[sort&compress]{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage{amsopn}
\usepackage{rotating}

@@

We developed a permutation-based procedure to factor out the effects of
temporal clustering from the clustering scores for each feature dimension. For
a given set of recalled items (whose presentation positions are given by $x_1,
x_2, x_3, …, x_N$), we circularly shift the presentation positions by a
randomly chosen amount (between 1 and the list length) to obtain a new set of
items. Since the new set of items will have the same (average) temporal
distances between successive recalls, the temporal clustering score for the new
set of items is equal (on average) to the temporal clustering score for the
original recalls. However, we can then re-compute the feature clustering score
for those new items. Finally, we can compute a ``temporally corrected'' feature
clustering score by computing the average percentile rank of the observed (raw)
feature clustering score within the distributions of circularly shifted feature
clustering scores, across $N = 500$ repetitions of this procedure. This new
temporally corrected score provides an estimate of the observed degree of
feature clustering over and above what could be accounted for by temporal
clustering alone.

@@

= 0.067$; Fig.~\ref{fig:barplots}A). Participants' performances on early lists
in all of the other order manipulation conditions were indistinguishable from
performance on the early feature rich lists ($\|t\|$s $< 1.013, p$s $> 0.314$).
Participants in both of the semantically ordered conditions exhibited stronger
temporal clustering on early lists (versus early feature rich lists; category:
$t(95) = 8.508, p < 0.001$; size: $t(95) = 2.429, p = 0.017$;
Fig.~\ref{fig:barplots}B). Participants in the length-ordered condition tended
to exhibit \textit{less} temporal clustering on early lists relative to early
feature rich lists ($t(95) = -1.666, p = 0.099$), whereas participants in the
first letter-ordered condition exhibited stronger temporal clustering on early
lists ($t(95) = 2.587, p = 0.011$). Participants in the visually ordered
conditions exhibited more similar performance on early lists, relative to early
feature rich lists (color: $t(96) = -1.064, p = 0.290$; we found a trending
enhancement for participants in the location-ordered condition: $t(95) = 1.682,
p = 0.096$). We also compared feature-based clustering on early lists across
the order manipulation and feature rich conditions. Since these results were
similar across both semantic conditions (category and size), both lexicographic
conditions (length and first letter), and both visual conditions (color and
location), here we aggregate data from conditions that manipulated each of
these three feature groupings in our comparisons, to simplify the presentation.
On early lists, participants in the semantically ordered conditions exhibited
stronger semantic clustering relative to participants in the feature rich
condition (category: $t(125) = 2.524, p = 0.013$; size: $t(125) = 3.510, p =
0.001$), but showed no reliable differences in lexicographic (length: $t(125) =
0.539, p = 0.591$; first letter: $t(125) = -0.587, p = 0.558$) or visual
(color: $t(125) = -0.579, p = 0.564$; location: $t(125) = -0.346, p = 0.730$)
clustering. Similarly, participants in the lexicographically ordered conditions
exhibited stronger (relative to feature rich participants) lexicographic
clustering (length: $t(125) = 3.426, p = 0.001$; first letter: $t(125) = 3.236,
p = 0.002$) on early lists, but showed no reliable differences in semantic
(category: $t(125) = -1.078, p = 0.283$; size: $t(125) = -0.310, p = 0.757$) or
visual (color: $t(125) = -0.209, p = 0.835$; location: $t(125) = -0.004, p =
0.997$) clustering. And participants in the visually ordered conditions
exhibited stronger visual clustering (again, relative to feature rich
participants, and on early lists; color: $t(126) = 2.099, p = 0.038$; location:
$t(126) = 4.392, p < 0.001$), but showed no reliable differences in semantic
(category: $t(126) = 0.204, p = 0.839$; size: $t(126) = -0.093, p = 0.926$) or
lexicographic (length: $t(126) = 0.714, p = 0.476$; first letter: $t(126) =
0.820, p = 0.414$) clustering. Taken together, these order manipulation results
suggest several broad patterns (Figs.~\ref{fig:recall-dynamics}A,
\ref{fig:fingerprints}). First, most of the order manipulations we carried out
did \textit{not} reliably affect overall recall performance. Second, most of
the order manipulations increased participants' tendencies to temporally
cluster their recalls. Third, all of the order manipulations enhanced
participants' clustering of each condition's target feature (i.e., semantic
manipulations enhanced semantic clustering, lexicographic manipulations
enhanced lexicographic clustering, and visual manipulations enhanced visual
clustering; Fig.~\ref{fig:barplots}C) while leaving clustering along other

@@

affect lexicographic or visual clustering, and so on). Although it is not
possible to fully separate feature versus temporal clustering when considering
sorted lists, we used a permutation-based procedure to identify the degree of
feature clustering over and above what could be accounted for by temporal
clustering alone (see \textit{Factoring out the effects of temporal
clustering}). When we carried out this analysis (Fig.~\ref{fig:barplots}D), we
found that participants exhibited more semantic clustering on semantically
sorted lists than on randomly ordered lists (category: $t(XXX) = XXX, p = XXX$;
size: $t(XXX) = XXX, p = XXX$), but the effects of the other order
manipulations could not reliably be separated from temporal clustering alone
($t$s $> XXX, p$s $> XXX$).

@@

\begin{sidewaysfigure}[p]

    \centering
    \includegraphics[width=\textwidth]{figures/memory_perf_barchart_compare}

\caption{\textbf{Recall probability and clustering scores on early and late
lists.} The bar heights display the average (across participants) recall
probabilities (\textbf{A.}), temporal clustering scores (\textbf{B.}), feature
clustering scores (\textbf{C.}), and residual feature clustering scores (after
factoring out temporal clustering effects; \textbf{D.}) for early (gray) and
late (gold) lists. For the feature rich bars (left), the feature clustering
scores are averaged across all feature dimensions. For the order manipulation
conditions, feature clustering scores are displayed for the focused-on feature
for each condition (e.g., category clustering scores are displayed for the
category condition, and so on). All panels: error bars denote
bootstrap-estimated 95\% confidence intervals. The horizontal dotted lines
denote the average values (across all lists and participants) for the feature
rich condition. The bars denote $t$-tests between the corresponding bars, and
the asterisks denote the Benjamini-Hochberg-corrected $p$-values. Comparisons
for which corrected $p \geq 0.05$ are not shown.} \label{fig:barplots}

\end{sidewaysfigure}
